From noise to knowledge: how randomness generates novel phenomena and information

### 2017-10-24 paper.Rmd
WIP

abstract: |
  In this paper, I provide a brief synthesis of ecological research in stochastic phenomena. Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath. Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.  Nor is all noise the same, and close examination of differences in in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.  While an exhaustive review is beyond the scope of this piece, I endeavor to highlight some of the most interesting examples of both noise-induced phenomena and noise-derived information, and point the way towards yet uncharted territory.

@@

```{r include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
library(ggthemes)
library(tidyverse)
library(cowplot)
library(hrbrthemes)
extrafont::loadfonts()
extrafont::font_import()
hrbrthemes::import_roboto_condensed()
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
#ggplot2::theme_set(theme_bw())
```

@@

Key questions

Key advances

@@

# Outline

## What is noise?

<!-- Our state variables don't "exist" except as averages of a lower level process -->

Where does noise come from?  It is important to bear in mind that whenever stochasticity is introduced into a model, it is only as a simplification or summary of the effects of a large number of other processes, often occuring at different scales.  Simply put, *noise is all that we leave out of our model*.  While ecologists will sometimes assert that stochasticity is inherent: birth and death are inherently stochastic events; or the environment is inherently stochastic, these are only assertions at a particular scale.  Obviously events such as births and deaths are themselves the results of mechanistic processes, just as fluctuations in rainfall and temperature are the result of deterministic laws of physics and not fundamentally random forces.  In all of science, only quantum mechanics claims a fundamentally probabilistic view of events, and then at such scales that have no relevance to ecology.  The term "noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually distinct blends together into a background of noise.

## Noise the nuisance: Noise can obscure patterns

@@

<!-- okay, claiming this for the last 2 decades might be overly provocative?-->

Over the past twenty years, ecological theory, modelling, and methods have increasingly elucidated the role of noise not only to *obscure* patterns but to *create* them.  This period marked a transistion away from the intense focus on simple models which exhibited deterministic chaos that had enchanted much of theoretical ecology in the decade prior in favor of more complex and stochastic models.  Of particular interest has been those conceptual advances that have bridged the divide from theoretical to empirical interest -- not merely the introduction of "real data" in a theoretical paper, but rather the concepts that take on a life of their own in the larger research community.

Our first stop will be into research into the origins of noise.  This will not only let us set up definitions for terms and concepts we will encounter throughout, but also introduce some of the fundamental mathematical and computational tooling needed in our tour of noisy phenemena.  From origins of noise, we will then explore some of the phenomena that noise can create.  These examples are central to the conceptual transistion away from viewing noise only as a nusiance that obscures patterns and realize the rich set of patterns and phenomena that can emerge.  Our tour will be by no means exhaustive, but as with chaos before it, theoretical allure does not always reflect ecological releveance.  Instead, we will pivot to our third

<!-- Origins -->

<!-- Phenomena -->

<!-- Applications -->

<!-- Connection to empiricists.  Not the same thing as connection to data. -->

## Origins of noise

## Sources of noise

@@

<!-- Demographic noise scaling with population size -->

<!-- Gillespie algorithm -->

@@

Colored noise e.g. [@Marshall2015]

@@

**Individual heterogeneity**

[@Hart2016]

**Spatial variability**

In addition to these sources of variation, there are additional sources of variation or uncertainty in a model which we will not consider for the purpose of this review which, while related, operate in a fundamentally different way that does not give rise to the same phenomena and information.

@@

- deterministic skeletons, 
- signal-to-noise ratio, 
- model fitting

## Noise the creator: Noise can induce novel phenomena

@@

```{r}
read_csv("appendix/quasicycles.csv") %>%
  gather(species, population, -t, -sigma) %>%
  rename(time = t) %>%
  ggplot(aes(time, population, col = species)) + geom_line() + 
  facet_wrap(~ sigma) + scale_color_ptol() #+ 
#  labs(x="Time", 
#       y="Population density",
#       title="Noise induces quasi-cycles",
#       caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
#ggsave("figures/quasicycles.pdf")
```

@@

## Noise the informer: Noise can provide novel information

@@

```{r}
tipping <- read_csv("appendix/tipping.csv")

p1 <- tipping %>%
  select(-x) %>%
  gather(series, value, -t, -tip_time) %>%
  ggplot(aes(t, value)) + geom_line() + 
  geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
  facet_wrap(~series, ncol = 2, scales="free_y") # +
#  theme_ipsum_rc() +
#  labs(title="A: Early warning signs")

p2 <- tipping %>% 
  select(t, population = x, tip_time) %>%
  ggplot(aes(t, population)) + geom_line() + 
  geom_vline(aes(xintercept = tip_time), col="red", lty=2) #+ 
#  scale_y_log10() # + 
#  theme_ipsum_rc() + 
#  labs(title="B: Population density",
#       caption = "Vertical red dashed line indicates tipping point location")

library(cowplot)
ggdraw() +
  draw_plot(p2, x = 0, y = 0, width = 1, height = 0.5) +
  draw_plot(p1, x = 0, y = .5, width = 1, height = 0.5)
#ggsave("figures/tipping.pdf")
```

@@

## Conclusions

## References

>>>>

layout: preprint
abstract: |
  In this paper, I provide a brief synthesis of ecological research in stochastic phenomena. Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath. Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.  Nor is all noise the same, and close examination of differences in in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.  While an exhaustive review is beyond the scope of this piece, I endeavor to highlight some of the most interesting examples of both noise-induced phenomena and noise-derived information, and point the way towards yet uncharted territory.

@@

```{r include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="CairoPDF", fig.width=3.5, fig.height=3.5)
library(ggthemes)
library(tidyverse)
library(cowplot)
library(hrbrthemes)
extrafont::loadfonts()
#extrafont::font_import()                    # One-time use to set up fonts
#hrbrthemes::import_roboto_condensed()       # One-time use to set up fonts
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
#ggplot2::theme_set(theme_bw()) # simpler theme if you do not have these fonts
```

@@

- Key questions
- Key advances

Moving beyond "everything is important" [@Bjornstad2001; @Kingsland1985]

@Bjornstad2001 future: more MCMC methods. More mechanistic models.  dealing with unobserved dynamics

@@

# Introduction: Noise the nuisance

@@

<!-- okay, claiming this for the last 2 decades might be overly provocative?-->
Over the past twenty years, ecological theory, modelling, and methods have increasingly elucidated the role of noise not only to *obscure* patterns but to *create* them.  This period marked a transistion away from the intense focus on simple models which exhibited deterministic chaos that had enchanted much of theoretical ecology in the previous two decades [@Bjornstad2001] in favor of more complex and stochastic models.  Of particular interest has been those conceptual advances that have bridged the divide from theoretical to empirical interest -- not merely the introduction of "real data" in a theoretical paper, but rather the concepts that take on a life of their own in the larger research community.  Consequently, this paper seeks to be a bridge between two audiences: theoretical researchers already expert in noisy phenomena but in interested in reaching wider empirical application, and also primarily empirical researchers seeking an entrance into noisy phenomena.

<!-- Regarding simulation examples -->
In this spirit, each of the examples illustrated here is coupled with consise and efficent R code in the appendix to both reproduce and further explore the simulations.  While examples of noisy phenomena in natural systems are also crucial, matching the pattern does not necessarily mean we have also matched the proccess behind the data.  The examples cited from the literature here offer some of the best empirical evidence of these phenomena, but I encourage readers to also experiment with the example code provided.  Simulations permit poking and prodding of empirical investigation quickly and with little mathematical formalism in the way.  In the code provided, most stochastic models are expressed in the BUGS language, which may be more familiar to empirical readers than corresponding mathematical formulae. This also permits both efficient simulation and potential estimation of parameters given sample data using the R package, NIMBLE [@DeValpine2017]. A copy of this appendix is maintained at <https://github.com/cboettig/noise-phenomena>, and bug reports, suggestions or help requests are welcome through the issue tracker.  A second appendix provides more mathematical details for interested readers.  With these in place, we are ready to begin our tour through three major areas of noisy phenomena: Origins of noise, emergent phenomena, and noise-driven inference.

<!-- MAP -->
Our first stop will be into research into the origins of noise.  This will not only let us set up definitions for terms and concepts we will encounter throughout, but also introduce some of the fundamental mathematical and computational tooling needed in our tour of noisy phenemena.  From origins of noise, we will then explore some of the phenomena that noise can create.  These examples are central to the conceptual transistion away from viewing noise only as a nusiance that obscures patterns and realize the rich set of patterns and phenomena that can emerge.  Our tour will be by no means exhaustive, but as with chaos before it, theoretical allure does not always reflect ecological releveance.  Having looked at some of the phenomena that can be created by noise, our third stop will try to run this in reverse: using noise to infer underyling phenomena.  Examples here are rarer but potentially the most rewarding, bridging the gap from theoretical construct to empirical investigation.  The most compelling examples have come from ways in which noisy pheneomena are used to infer stability and forecast changes in resilience of populations or ecosystems. Though not without weaknesses or missteps, the resulting dialog between theoretical and empirical aspects have made this particularly a rich area of research.  We end our tour by considering what features best enable the translation and ecological relevance of noisy phenomena.

# Origins of noise

What is noise? Where does it come from?  It is important to bear in mind that whenever stochasticity is introduced into a model, it is only as a simplification or summary of the effects of a large number of other processes, often occuring at different scales.  Simply put, *noise is all that we leave out of our model*.  While ecologists will sometimes assert that stochasticity is inherent: birth and death are inherently stochastic events; or the environment is inherently stochastic, these are only assertions at a particular scale.  Obviously events such as births and deaths are themselves the results of mechanistic processes, just as fluctuations in rainfall and temperature are the result of deterministic laws of physics and not fundamentally random forces.  In all of science, only quantum mechanics claims a fundamentally probabilistic view of events, and then at such scales that have no relevance to ecology.  The term "noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually distinct blends together into a background of noise.  
<!-- Our state variables don't "exist" except as averages of a lower level process -->

@@

<!-- Gillespie algorithm; Population level (SDE) vs agent-based vs individual  -->

@@

<!-- Demographic noise scaling away with population size, but not so for environmental noise -->

[@Bjornstad2001] _Science_ Review of noisy fluctuations

[@Melbourne2008] Importance of demographic stochasticity and its interaction with environmental stochasticity in driving extinction risk. 
[@Bonsall2004] Markov chain environmental and demographic noise predator-prey metapopulation model.

Colored noise:
- [@Ripa1999]
- [@Marshall2015]

@@

<!-- non-temporal stochasticity -->

**Other forms of stochasticity**.  Stochasticity frequently enters our models in other ways as well.  For instance, most population-level models treat all individuals within the population, or at least within a particular age or stage class, as identical.  *Individual heterogeneity* in relevant traits (growth rates, dispersal ability) can be represented probabilistically as a distribution rather than a fixed value, making it function as another form of random varation in a model.  *Spatial heterogeneity* can likewise be reflected in stochastic parameters of a spatially implicit model, or as spatially varying environmental stochasticity in a spatially explicit model.  As with other forms of noise [@Melbourne2008; @Bonsall2004], interaction between individual or spatial variation and other sources of noise can drive patterns including the potential for persistence, coexistence, and dispersal of species [@Schreiber2010; @Hart2016; @Schreiber2017], but are beyond the scope of this paper.

@@

# Noise the creator: Noise can induce novel phenomena

@@

```{r fig.width=7}
read_csv("appendix/quasicycles.csv") %>%
  gather(species, population, -t, -sigma) %>%
  rename(time = t) %>%
  ggplot(aes(time, population, col = species)) + geom_line() + 
  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() #+ 
#  labs(x="Time", 
#       y="Population density",
#       title="Noise induces quasi-cycles",
#       caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
#ggsave("figures/quasicycles.pdf")
```

@@

# Noise the informer: Noise can provide novel information

@@

```{r fig.width=7, fig.height=7}
tipping <- read_csv("appendix/tipping.csv")

p1 <- tipping %>%
  select(-x) %>%
  gather(series, value, -t, -tip_time) %>%
  ggplot(aes(t, value)) + geom_line() + 
  geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
  facet_wrap(~series, ncol = 2, scales="free_y")  +
  theme_ipsum_rc() 
p2 <- tipping %>% 
  select(t, population = x, tip_time) %>%
  ggplot(aes(t, population)) + geom_line() + 
  geom_vline(aes(xintercept = tip_time), col="red", lty=2) + 
  #scale_y_log10()  + 
  theme_ipsum_rc() + 
  labs(subtitle = "Yeast population size", 
       caption = "Vertical red dashed line indicates tipping point location")

library(cowplot)
ggdraw() +
  draw_plot(p1, x = 0, y = .5, width = 1, height = 0.5) +
  draw_plot(p2, x = 0, y = 0, width = 1, height = 0.6)
#ggsave("figures/tipping.pdf")
```

@@

# Conclusions

# References
